On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2023-10-12 Manuscript/manuscript.tex
Aligned something
Going over Prof Yan's comments

To test the conjecture that the 2022 World Championships timing device may have 
led to faster recorded reaction times, we compare the reaction times of the same
athletes who have attended both the 2022 World Championships and other 
competitions. 
In this setting, we have clustered data with subunit grouping. In particular,
each athlete is a cluster and the multiple reaction times from the same athlete
can be from either the 2022 World Championships or otherwise.
Let $X_{ij}$ be the $j$th reaction time of athlete~$i$, $i = 1, \ldots, n$,
$j = 1, \ldots, m_i$ where $m_i$ is the number of observations from
athelete~$i$. Let $\delta_{ij}$ be the group indicator of $X_{ij}$; $\delta_{ij}
= 1$ if $X_{ij}$ is in group~1 (2022 World Championships) and $\delta_{ij} = 0$ 
otherwise. Athletes are
assumed to be independent, while subunit observations from the same athlete are
not. The null hypothesis $H_0$ to be tested is that there is no difference
between the two groups; i.e., the distribution of $X_{ij}$ remains the same
regardless of the group indicator $\delta_{ij}$.

@@

\subsection{GLMM}\label{sec:glmm}
Based on an exploratory analysis using the GLMM model for the reaction times
from the World Championships, the we chose a gamma mixed-effect model as our 
final model. As suggested by the log-likelihood and the Akaike Information
Criterion (AIC), the normal error model on the log-scale was found to be
much inferior to the gamma model which better captures the skewness of the
data. Two random effects were found useful: one capturing the venue effect in
each year and the other capturing the heat effect within each year. The heat
effect is important as it captures the variability in the amount of time
athletes are on the starting blocks before the gun goes off. Thus, we can try
to reduce this variability by accounting for it in the model through the heat
effect.

@@

\jy{Add the results without 2022 to this table.}
\begin{table}
  \centering
  \caption{AIC and estimated parameters for the three models fitted to all the
    data.}
  \label{tab:Gamma_parameters}
  \begin{tabular}{c c c c c c}
      \toprule
      Model & AIC & $\sigma_h$ & $\sigma_v$ & $\alpha$ & $\sqrt{\phi}$ \\
      \midrule
      Venue Effect Only & $-2023.5$ &   & 0.054 & $-1.871$ & 0.150\\
      Heat Effect Only & $-2070.7$ & 0.073 &  & $-1.890$ & 0.135\\
      Venue and Heat Effect & $-2075.7$ & 0.057 & 0.040 & $-1.880$ & 0.134\\
      \bottomrule
  \end{tabular}
\end{table}

The estimated parameters of the GLMM model from our analysis are shown in 
Table~\ref{tab:Gamma_parameters}. These statistics reflect the inclusion of 2022
in our analysis and shows the three models we considered. The standard deviation
of the heat effect is $\sigma_h$, the standard deviation of the venue effect
is $\sigma_v$, the intercept is $\alpha$ and we report $\sqrt{\phi}$ as the
square root of the dispersion parameter $\phi$.
The GLMM with both venue and heat random effects was the best model as evidenced
by a decrease in the AIC compared to the other models with only one random
effect.
\jy{discuss the differences in the parameter estimates with and without 2022
  data, especially the sd of the random effect of venue.}

>>>>

To test the conjecture that the 2022 World Championships timing device may have 
led to faster recorded reaction times, we compare the reaction times of the same
athletes who have attended both the 2022 World Championships and other 
competitions. 
In this setting, we have clustered data with subunit grouping. In particular,
each athlete is a cluster and the multiple reaction times from the same athlete
can be from either the 2022 World Championships or otherwise.
Let $X_{ij}$ be the $j$th reaction time of athlete~$i$, $i = 1, \ldots, n$,
$j = 1, \ldots, m_i$ where $m_i$ is the number of observations from
athlete~$i$. Let $\delta_{ij}$ be the group indicator of $X_{ij}$; $\delta_{ij}
= 1$ if $X_{ij}$ is in group~1 (2022 World Championships) and $\delta_{ij} = 0$ 
otherwise. Athletes are
assumed to be independent, while subunit observations from the same athlete are
not. The null hypothesis $H_0$ to be tested is that there is no difference
between the two groups; i.e., the distribution of $X_{ij}$ remains the same
regardless of the group indicator $\delta_{ij}$.

@@

\subsection{GLMM}\label{sec:glmm}
Based on an exploratory analysis using the GLMM model for the reaction times
from the World Championships, we chose a gamma mixed-effect model as our 
final model. As suggested by the log-likelihood and the Akaike Information
Criterion (AIC), the normal error model on the log-scale was found to be
much inferior to the gamma model which better captures the skewness of the
data. Two random effects were found useful: one capturing the venue effect in
each year and the other capturing the heat effect within each year. The heat
effect is important as it captures the variability in the amount of time
athletes are on the starting blocks before the gun goes off. Thus, we can try
to reduce this variability by accounting for it in the model through the heat
effect.

@@

\jy{Add the results without 2022 to this table.}
\of{Does it make sense to order the table how I did, and block by dataset or
should I block by model?}
\begin{table}
  \centering
  \caption{AIC and estimated parameters for the three models fitted to all the
    data.}
  \label{tab:Gamma_parameters}
  \begin{tabular}{c c c c c c}
      \toprule
      Model & AIC & $\sigma_h$ & $\sigma_v$ & $\alpha$ & $\sqrt{\phi}$ \\
      \midrule
      With 2022 Venue Effect Only & $-2023.5$ &   & 0.054 & $-1.871$ & 0.150\\
      With 2022 Heat Effect Only & $-2070.7$ & 0.073 &  & $-1.890$ & 0.135\\
      With 2022 Venue and Heat Effect & $-2075.7$ & 0.057 & 0.040 & $-1.880$ & 0.134\\
      \midrule
      Without 2022 Venue Effect Only & $-1862.2$ &   & 0.048 & $-1.857$ & 0.152\\
      Without 2022 Heat Effect Only & $-1910.9$ & 0.069 &  & $-1.877$ & 0.136\\
      Without 2022 Venue and Heat Effect & $-1912.9$ & 0.058 & 0.033 & $-1.880$ & 0.136\\
      \bottomrule
  \end{tabular}
\end{table}

The estimated parameters of the GLMM model from our analysis are shown in 
Table~\ref{tab:Gamma_parameters}. These statistics reflect the inclusion of 2022
in our analysis and shows the three models we considered. The standard deviation
of the heat effect is $\sigma_h$, the standard deviation of the venue effect
is $\sigma_v$, the intercept is $\alpha$ and we report $\sqrt{\phi}$ as the
square root of the dispersion parameter $\phi$.
The GLMM with both venue and heat random effects was the best model as evidenced
by a decrease in the AIC compared to the other models with only one random
effect.  When we exclude 2022 from the full model, we observe small but intriguing
discrepancies in the models. The standard deviation of the venue effect, 
$\sigma_v$, decreases from 0.40 to 0.33 when 2022 is excluded.  Thus, 2022
increased the noise in the overall data and confirms that it is an outlier.

\jy{discuss the differences in the parameter estimates with and without 2022
  data, especially the sd of the random effect of venue.}
